---
title: "Survey of National Oceanic & Atmospheric Administration Storm Data"
author: "Al Kepner"
date: "2/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Survey of Storm Event Data Set from NOAA 1950-2011
## Coursera Project 2 for Reproducible Research by Al Kepner


# Synopsis
TBD

# Processing

### Load libraries and load the data set
### Download the data file into the ./data directory
### Load the zipped CSV file into a data frame

```{r, cache=TRUE}
library(dplyr)
library(ggplot2)
library(reshape2)
library(lubridate)
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipFile <- "data/StormData.csv.bz2"
download.file(url=url,destfile=zipFile,method="auto")
## Load the zipped CSV file into a data frame
zipFile <- "data/StormData.csv.bz2"
stormData1 <- read.csv(zipFile)

```

### For clarity, select only the columns relating to Date, location(state)
### Event Type, harm to persons, economic impact
```{r, cache=TRUE}
StormFocus1 <- stormData1 %>% 
  select(BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG, PROPDMGEXP, CROPDMGEXP)
```

### For harm to persons we will sum the FATALITIES and INJURIES columns.
###
### For economic impact we will sum the PROPDMG and CROPDMG columns,
### after adjusting to use the appropriate scale factor or
### K -- Kilo, M -- Millions, or B -- Billions from the corresponding 
### PROPDMGEXP and CROPDMGEXP columns.
##3
### For time of event we will just extract the year from BGN_DATE


### Define a function to multiply a damage cost column (PROPDMG or CROPDMG)
### by the appropriate scale factor.
```{r, cache=TRUE}
mapCost <- function(cost, mult) {
  if(mult == "b" || mult == "B") {
    result <- cost * 1e9
  } else if(mult == "m" || mult == "M") {
    result <- cost * 1e6
  } else if(mult == "k" || mult == "K") {
    result <- cost * 1e3
  } else {
    result <- cost
  }
  result
}
```

#### This transforms the BGN_DATE in format "m/d/y hh:mm:ss" to just the year as an integer.
```{r, cache=TRUE}
year <- as.integer(sub(" 0:00:00","",sub("[0-9]+/[0-9]+/","",as.character(StormFocus1$BGN_DATE))))
StormFocus2 <- cbind(StormFocus1,year)
summary(StormFocus2)
```

#### We use the mapCost function defined above to calculate property_cost and
#### crop_cost columns. We sum total storm damage cost as total_cost for these 2 
#### columns. We also sum FATALITIES plus INJURIES as harm_to_persons.
```{r, cache=TRUE}
StormFocus3 <- StormFocus2 %>% rowwise() %>%
  mutate(property_cost=mapCost(PROPDMG, PROPDMGEXP), 
         crop_cost=mapCost(CROPDMG, CROPDMGEXP),
         total_cost=sum(property_cost, crop_cost, na.rm=TRUE),
         harm_to_persons=sum(FATALITIES,INJURIES, na.rm=TRUE)) %>%
           ungroup()
```

#### Here we select just columns of interest for ranking types of
#### storm events.
```{r, cache=TRUE}
StormFocus4 <- StormFocus3 %>% rename(Event_Type = EVTYPE) %>%  
  select(STATE, year, Event_Type, FATALITIES, INJURIES, harm_to_persons, 
         property_cost, crop_cost, total_cost)
summary(StormFocus4)
```


### Rank Event_Type by total_cost descending (summed across all years and locations)
```{r, cache=TRUE}
TotalCostRank <- StormFocus4 %>% group_by(Event_Type) %>%
  summarize(TotalCost = sum(total_cost), Count=n()) %>%
  arrange(desc(TotalCost))

TotalCostRank
```



### Rank Event_type by harm_to_persons descending(summed across all years and locations)
```{r, cache=TRUE}
HarmToPersonsRank <- StormFocus4 %>% group_by(Event_Type) %>%
  summarize(HarmToPersons = sum(harm_to_persons), Count=n()) %>%
  arrange(desc(HarmToPersons))

HarmToPersonsRank
```


