---
title: "Survey of National Oceanic & Atmospheric Administration Storm Data"
author: "Al Kepner"
date: "2/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Survey of Storm Event Data Set from NOAA 1950-2011
## Coursera Project 2 for Reproducible Research by Al Kepner


# Synopsis
This report explores the total cost of property damage and crop damage categorized 
by weather event types. The event types are ranked by total cost over all data for 
years 1950-2011 to determine types of events causing the most economic damage.
We also rank the event types by the sum of fatalities and injuries to determine which 
types of events cause the most harm to persons.

# Data Processing

Load libraries and load the data set.  
Download the data file into the ./data directory.  
Load the zipped CSV file into a data frame.  

```{r, cache=TRUE}

library(dplyr)
library(ggplot2)
library(reshape2)
library(lubridate)

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipFile <- "data/StormData.csv.bz2"
download.file(url=url,destfile=zipFile,method="auto")
## Load the zipped CSV file into a data frame
zipFile <- "data/StormData.csv.bz2"
stormData1 <- read.csv(zipFile)

```

For clarity, select only the columns relating to Date, location(state)
Event Type, harm to persons, economic impact
```{r, cache=TRUE}
StormFocus1 <- stormData1 %>% 
  select(BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG, PROPDMGEXP, CROPDMGEXP)
```

For harm to persons we will sum the FATALITIES and INJURIES columns.

For economic impact we will sum the PROPDMG and CROPDMG columns,
after adjusting to use the appropriate scale factor or
K -- Kilo, M -- Millions, or B -- Billions from the corresponding 
PROPDMGEXP and CROPDMGEXP columns.

For time of event we will just extract the year from BGN_DATE


Define a function to multiply a damage cost column (PROPDMG or CROPDMG)
by the appropriate scale factor.
```{r, cache=TRUE}
mapCost <- function(cost, mult) {
  if(mult == "b" || mult == "B") {
    result <- cost * 1e9
  } else if(mult == "m" || mult == "M") {
    result <- cost * 1e6
  } else if(mult == "k" || mult == "K") {
    result <- cost * 1e3
  } else {
    result <- cost
  }
  result
}
```

#### This transforms the BGN_DATE in format "m/d/y hh:mm:ss" to just the year as an integer.
```{r, cache=TRUE}
year <- as.integer(sub(" 0:00:00","",sub("[0-9]+/[0-9]+/","",as.character(StormFocus1$BGN_DATE))))
StormFocus2 <- cbind(StormFocus1,year)
summary(StormFocus2)
```

#### We use the mapCost function defined above to calculate property_cost and
#### crop_cost columns. We sum total storm damage cost as total_cost for these 2 
#### columns. We also sum FATALITIES plus INJURIES as harm_to_persons.
```{r, cache=TRUE}

StormFocus3 <- StormFocus2 %>% rowwise() %>%
  mutate(property_cost=mapCost(PROPDMG, PROPDMGEXP), 
         crop_cost=mapCost(CROPDMG, CROPDMGEXP),
         total_cost=sum(property_cost, crop_cost, na.rm=TRUE),
         harm_to_persons=sum(FATALITIES,INJURIES, na.rm=TRUE)) %>%
           ungroup()
```

#### Here we select just columns of interest for ranking types of
#### storm events.
```{r, cache=TRUE}
StormFocus4 <- StormFocus3 %>% rename(Event_Type = EVTYPE) %>%  
  select(STATE, year, Event_Type, FATALITIES, INJURIES, harm_to_persons, 
         property_cost, crop_cost, total_cost)
summary(StormFocus4)
```


#### Rank Event_Type by total_cost descending (summed across all years and locations)
```{r, cache=TRUE}
TotalCostRank <- StormFocus4 %>% group_by(Event_Type) %>%
  summarize(TotalCost = sum(total_cost), Count=n()) %>%
  arrange(desc(TotalCost))

TotalCostRank
```



#### Rank Event_type by harm_to_persons descending(summed across all years and locations)
```{r, cache=TRUE}
HarmToPersonsRank <- StormFocus4 %>% group_by(Event_Type) %>%
  summarize(HarmToPersons = sum(harm_to_persons), Count=n()) %>%
  arrange(desc(HarmToPersons))

HarmToPersonsRank
```

We have found that TORNADO events been the greatest contributor to
Harm to persons in aggregate over the the entire period.
We would now like to construct a plot which shows how TORNADO related
FATALITIES and INJURIES have varied by year.
```{r}
summary(StormFocus3)
```
StormFocus3 has a selected subset of columns of all observations from the original data set,
and includes the columns we want, YEAR, FATALITIES, INJURIES. We
want to group and summarize this data by year.




```{r}
TornadoHarmToPersonsByYear <- StormFocus3 %>%
  filter(EVTYPE == "TORNADO") %>%
  group_by(year) %>%
  summarize(Fatalities=sum(FATALITIES),Injuries=sum(INJURIES))
```


We also want to summarize costs by year for FLOOD damage,
which is the highest overall cost event type.
```{r}
FloodDamageCostByYear <- StormFocus3 %>%
  filter(EVTYPE == "FLOOD") %>%
  group_by(year) %>%
  summarize(Crop_Damage=sum(crop_cost/1e6),
            Property_Damage=sum(property_cost/1e6))
```

```{r}
fdt <- data.frame(FloodDamageCostByYear)
fdt
```

It appears that FLOOD was not recorded as an Event Type before 1993.
This query confirms the count of rows per year for EVTYPE FLOOD.
```{r}
FloodCount <- StormFocus3 %>%
  filter(EVTYPE == "FLOOD") %>%
  group_by(year) %>% summarize(count = n())
head(FloodCount)
range(FloodCount$year)
```

The following plot shows a timeline of how Tornado related 
deaths and injuries have varied by year
```{r}
th <- TornadoHarmToPersonsByYear
df <- data.frame(Year=th$year, 
                 Fatalities=th$Fatalities,
                 Injuries=th$Injuries)
str(df)
df2 <- melt(df, id.vars='Year', variable.name='Event', value.name='Count')
str(df2)

g <- ggplot(df2, aes(Year,Count)) + geom_line(aes(color=Event)) +
  labs(title="Tornado Deaths and Injuries by Year 1950-2011")
g
```

The following plot shows a timeline of how FLOOD damage
costs have varied by year.
```{r}
fd <- FloodDamageCostByYear
df <- data.frame(Year=fd$year,
                 Crops=fd$Crop_Damage,
                 Property=fd$Property_Damage)
df2 <- melt(df, id.vars="Year", variable.name="DamageType", value.name='Damage')
str(df2)

g <- ggplot(df2, aes(Year,Damage)) +
  geom_line() +
  facet_grid(DamageType ~ .) +
  labs(title="Flood Damage to Crops and Property by Year 1993-2011") +
  ylab('Damage $ in millions')
g
```

It appears that there is an outlier in FLOOD damage cost for 2006
which dominates the other data.
We can dig into this a little.
Find maximum property damage row from original data set.
```{r}
hicost <- max(StormFocus3$property_cost)

hicost/1e9

row <- StormFocus3[StormFocus3$property_cost == hicost,]

row
```
There was apparently one flood event in CA in 2006 which
which was recorded at 115 Billion in property damage.
This one event dominates the other data on weather damage costs.
The total FLOOD damage recorded in this data set is about
150 billion for all years.

# Results

- In the above analysis, the variable TotalCostRank shows that the top 3 Event Types
for economic damage were FLOOD, HURRICANE/TYPHOON, and TORNADO, with FLOOD showing
a total of about $150 billion damage over the entire data set.

- The variable HarmToPersonsRank shows that the top event types causing fatalities and injuries
were TORNADO, EXCESSIVE HEAT, and TSTM WIND.

- We also showed timelines for FLOOD caused economic damage, and TORNADO harm to persons.


