---
title: "Survey of National Oceanic & Atmospheric Administration Storm Data"
author: "Al Kepner"
date: "2/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Survey of Storm Event Data Set from NOAA 1950-2011
## Coursera Project 2 for Reproducible Research by Al Kepner


# Synopsis
TBD

# Processing

## Load libraries and load the data set
## Download the data file into the ./data directory
## Load the zipped CSV file into a data frame

```{r, cache=TRUE}
library(dplyr)
library(ggplot2)
library(reshape2)
library(lubridate)
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipFile <- "data/StormData.csv.bz2"
download.file(url=url,destfile=zipFile,method="auto")
## Load the zipped CSV file into a data frame
zipFile <- "data/StormData.csv.bz2"
stormData1 <- read.csv(zipFile)

```

## For clarity, select only the columns relating to Date, location(state)
## Event Type, harm to persons, economic impact
```{r, cache=TRUE}
StormFocus1 <- stormData1 %>% 
  select(BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG, PROPDMGEXP, CROPDMGEXP)
```

## For harm to persons we will sum the FATALITIES and INJURIES columns.
##
## For economic impact we will sum the PROPDMG and CROPDMG columns,
## after adjusting to use the appropriate scale factor or
## K -- Kilo, M -- Millions, or B -- Billions from the corresponding 
## PROPDMGEXP and CROPDMGEXP columns.
##
## For time of event we will just extract the year from BGN_DATE


## Define a function to multiply a damage cost column (PROPDMG or CROPDMG)
## by the appropriate scale factor.
```{r, cache=TRUE}
mapCost <- function(cost, mult) {
  if(mult == "b" || mult == "B") {
    result <- cost * 1e9
  } else if(mult == "m" || mult == "M") {
    result <- cost * 1e6
  } else if(mult == "k" || mult == "K") {
    result <- cost * 1e3
  } else {
    result <- cost
  }
  result
}
```

### This transforms the BGN_DATE in format "m/d/y hh:mm:ss" to just the year as an integer.
```{r, cache=TRUE}
year <- as.integer(sub(" 0:00:00","",sub("[0-9]+/[0-9]+/","",as.character(StormFocus1$BGN_DATE))))
StormFocus2 <- cbind(StormFocus1,year)
summary(StormFocus2)
```

